---
title: "Performance benchmark: heatwave3 vs heatwaveR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance benchmark: heatwave3 vs heatwaveR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

This vignette benchmarks heatwave3 against heatwaveR on a realistic gridded
SST dataset, the full OSTIA South-West Indian Ocean reanalysis (400 lon × 200
lat, ca. 49,800 ocean pixels, 14,276 daily time steps, 1982–2021).

We compare three configurations:

1. **heatwaveR**, serial per-pixel R + C++ (the standard approach)
2. **heatwave3 (1 thread)**, pure C++ with single-threaded execution
3. **heatwave3 (12 threads)**, pure C++ with OpenMP parallelism

## Test data

```{r data}
sst_file <- "/Volumes/OceanData/OSTIA_East_Coast_MHW/SWIO_Jan1982-Dec2021.nc"
# 400 lon × 200 lat at 0.05° resolution
# 15–35°E, 38–28°S (Agulhas Current system)
# 14,276 daily time steps (1982-01-01 to 2021-12-31)
# ca. 49,796 ocean pixels (remainder is land)
```

## heatwave3: single-threaded

```{r hw3-1t}
library(heatwave3)

stem <- file.path(tempdir(), "bench")

t_clim <- system.time(
  ts2clm3(sst_file, name = stem,
          climatologyPeriod = c("1982-01-01", "2011-12-31"),
          var_name = "analysed_sst", n_threads = 1L, quiet = TRUE)
)

t_event <- system.time(
  detect_event3(sst_file, name = stem,
                var_name = "analysed_sst", n_threads = 1L, quiet = TRUE)
)

cat("Climatology:", round(t_clim[3], 1), "sec\n")
cat("Detection:  ", round(t_event[3], 1), "sec\n")
cat("Total:      ", round(t_clim[3] + t_event[3], 1), "sec\n")
```

## heatwave3: 12 threads

```{r hw3-12t}
t_clim_12 <- system.time(
  ts2clm3(sst_file, name = stem,
          climatologyPeriod = c("1982-01-01", "2011-12-31"),
          var_name = "analysed_sst", n_threads = 12L, quiet = TRUE)
)

t_event_12 <- system.time(
  detect_event3(sst_file, name = stem,
                var_name = "analysed_sst", n_threads = 12L, quiet = TRUE)
)

cat("Climatology:", round(t_clim_12[3], 1), "sec\n")
cat("Detection:  ", round(t_event_12[3], 1), "sec\n")
cat("Total:      ", round(t_clim_12[3] + t_event_12[3], 1), "sec\n")
```

## heatwaveR: serial baseline

heatwaveR processes one pixel at a time. We time a 20-pixel sample (data
pre-loaded to isolate computation from I/O) and extrapolate to the full grid.

```{r heatwaveR}
library(heatwaveR)
library(ncdf4)

nc <- nc_open(sst_file)
time_raw <- ncvar_get(nc, "time")
dates <- as.Date("1981-01-01") + time_raw / 86400

# Pre-load one longitude column (200 lat pixels)
sst_col <- ncvar_get(nc, "analysed_sst",
                     start = c(200, 1, 1), count = c(1, 200, -1))
nc_close(nc)

ocean_idx <- which(!is.na(sst_col[, 1]))
n_test <- 20L

t_hwr <- system.time({
  for (j in ocean_idx[1:n_test]) {
    clm <- ts2clm(data.frame(t = dates, temp = as.numeric(sst_col[j, ])),
                  climatologyPeriod = c("1982-01-01", "2011-12-31"))
    ev <- detect_event(clm)
  }
})

per_pixel <- t_hwr[3] / n_test
n_ocean <- 49796  # pre-counted
estimated <- per_pixel * n_ocean

cat("Per pixel:      ", round(per_pixel, 3), "sec\n")
cat("Estimated total:", round(estimated), "sec (",
    round(estimated / 60, 1), "min)\n")
```

## Results

Benchmarked on an Apple M3 Pro (12-core), macOS 15.5, R 4.5.3.

| Method | Climatology | Detection | Total | Speedup |
|--------|------------|-----------|-------|---------|
| heatwaveR (serial) | n/a | n/a | ca. 4,116 sec (68.7 min) | 1× |
| heatwave3 (1 thread) | 174 sec | 82 sec | **257 sec (4.3 min)** | **16×** |
| heatwave3 (12 threads) | 60 sec | 55 sec | **115 sec (1.9 min)** | **36×** |

The speedup comes from three sources:

1. **C++ algorithms**. Climatology (sliding-window percentile) and event
   detection (run-length encoding, gap joining, 19 metrics) are implemented
   in pure C++ rather than R + data.table.
2. **OpenMP parallelism**. Each pixel is independent, so OpenMP distributes
   pixels across cores with zero coordination overhead.
3. **Direct NetCDF I/O**. Reads the full data slab in one C call via
   libnetcdf, avoiding R's per-pixel ncdf4 random-access overhead.

### Scaling notes

- The 12-thread run is 2.2× faster than single-threaded, not 12×, because the
  full 400×200 grid must first be read into memory (ca. 900 MB for this dataset)
  and the NetCDF I/O is single-threaded. The parallelism benefit is in the
  computation phase only.
- For grids that fit in memory, I/O accounts for roughly 40–60% of wall time.
  Larger grids with more ocean pixels show better parallel scaling as the
  compute fraction grows.
- On Linux with gcc OpenMP (which has lower thread-creation overhead than macOS
  libomp), parallel scaling is typically better.

## Benguela region: daily files, full pipeline

This benchmark applies heatwave3 to a production-scale analysis, the
Benguela upwelling system, using **16,049 individual daily OSTIA files**
(1982–2025) with the WMO 1991–2020 climatological baseline.

The subsetted OSTIA files cover 7–20°E, 35–17°S at 0.05° resolution (260 lon ×
360 lat = 93,600 pixels). This is a region with strong upwelling variability and
a mix of ocean and land pixels.

```{r benguela-setup, eval = FALSE}
library(heatwave3)

ostia_dir <- "/Volumes/OceanData/Tom/OSTIA"
stem <- file.path(tempdir(), "benguela")
```

```{r benguela-clim, eval = FALSE}
# Climatology from ca. 16,000 daily files (WMO 1991-2020 baseline)
t_clim <- system.time(
  ts2clm3(
    file_in  = ostia_dir,
    name     = stem,
    climatologyPeriod = c("1991-01-01", "2020-12-31"),
    var_name = "analysed_sst",
    n_threads = 12L
  )
)
```

```{r benguela-detect, eval = FALSE}
# Event detection on the same daily files
t_event <- system.time(
  detect_event3(
    file_in   = ostia_dir,
    name      = stem,
    var_name  = "analysed_sst",
    n_threads = 12L
  )
)
```

### Benguela results

| Step | Time | Detail |
|------|------|--------|
| Read + merge 16,049 daily files | included | sorted, deduplicated, assembled into 3D cube |
| Climatology (12 threads) | **95 sec** | 30-year WMO baseline, 93,600 pixels × 366 DOYs |
| Event detection (12 threads) | **108 sec** | threshold exceedance, RLE, gap joining, 19 metrics |
| **Total** | **203 sec (3.4 min)** | 8.4 million events detected |

For comparison, processing 93,600 pixels serially with heatwaveR at ca. 0.083
sec/pixel would take ca. **2.2 hours**. heatwave3 completes the same
analysis in 3.4 minutes, a **ca. 38× speedup**.
