---
title: "Performance benchmark: heatwave3 vs xmhw"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance benchmark: heatwave3 vs xmhw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Pretty output
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)

# Environment for python
# Sys.setenv(
#   LD_LIBRARY_PATH = paste(
#     "/home/calanus/anaconda3/envs/xmhw_env/lib",
#     Sys.getenv("LD_LIBRARY_PATH"),
#     sep = ":"
#   )
# )
```

## Overview

This vignette benchmarks `heatwave3` against the python module `xmhw` on the OSTIA gridded
SST dataset for the French coastline of the Mediterranean Sea 
(104 lon × 60 lat, 6,240 ocean pixels, 15,706 daily time steps, 1982–2024).

We compare three configurations:

1. **heatwaveR**, serial per-pixel R + C++ (the standard approach)
2. **heatwave3 (1 thread)**, pure C++ with single-threaded execution
3. **xmhw (1 thread)**, pure python without dask
4. **heatwave3 (12 threads)**, pure C++ with multi-threading
5. **xmhw (12 threads)**, pure python with dask enabled

```{r setup}
# Necessary R packages
library(reticulate)
library(ncdf4)
library(terra)
library(heatwaveR)
library(heatwave3)
```

```{r check-env, include=FALSE}
# cat(Sys.getenv("LD_LIBRARY_PATH"))
# reticulate::py_config()
```

## Test data

```{r data}
# Choose bounding box
bbox <- c(
  2.5,   # °E  (just west of Cerbère)
  41.0,  # °N  (south of Corsica, open sea)
  7.7,   # °E  (Menton / Ligurian coast)
  44.0   # °N  (northern fringe, pre-Alps shelf)
)

# Select daterange
time_range <- c(
  "1982-01-01",
  "2024-12-31"
)

# Download the desired OSTIA cube
# OSTIA_data <- CopernicusMarine::cms_download_subset(
#   product = "SST_GLO_SST_L4_REP_OBSERVATIONS_010_011",
#   layer = "METOFFICE-GLO-SST-L4-REP-OBS-SST",
#   variable = "analysed_sst",
#   region = bbox,
#   time = time_range,
#   username = "...", # Replace with your CMEMS username
#   password = "...", # Replace with your CMEMS password
#   progress = TRUE
# )

# Or simply download them via the UI
# https://data.marine.copernicus.eu/product/SST_GLO_SST_L4_REP_OBSERVATIONS_010_011/download?dataset=METOFFICE-GLO-SST-L4-REP-OBS-SST_202003

# The NetCDF file for further use
sst_file <- "/home/calanus/data/OSTIA/METOFFICE-GLO-SST-L4-REP-OBS-SST_French_Med.nc"
# 104 lon × 60 lat at 0.05° resolution
# 2.5–7.7°E, 41–44°N (French Mediterranean coastline)
# 15,706 daily time steps (1982-01-01 to 2024-12-31)
# 6,240 ocean pixels (remainder is land)
```

## Python environment

We will control the comparison across R and python by using an existing python v3.11 environemnt 
created for this tutorial with the following modules installed: `netCDF4`, `xarray`, `dask`, and `xmhw`.

```{r reticulate}
# Activate the conda environment and load functionality into R environment
use_condaenv("RiOMar", required = TRUE)
nc4 <- import("netCDF4")
xmhw <- import("xmhw.xmhw")
xr <- import("xarray")
dask <- import("dask")
builtins <- import_builtins()

# Pull out the two key functions
x_threshold <- xmhw$threshold
x_detect <- xmhw$detect
py_slice <- builtins$slice
```

## heatwaveR: serial baseline

`heatwaveR` processes one pixel at a time. We time a 20-pixel sample (data
pre-loaded to isolate computation from I/O) and extrapolate to the full grid.

```{r heatwaveR}
nc <- nc_open(sst_file)
time_raw <- ncvar_get(nc, "time")
dates <- as.Date("1970-01-01") + time_raw / 86400

# Pre-load one longitude column (20 lat pixels)
sst_col <- ncvar_get(nc, "analysed_sst",
                     start = c(1, 1, 1), 
                     count = c(1, 20, -1))
nc_close(nc)

ocean_idx <- which(!is.na(sst_col[, 1]))
n_test <- 20L

t_hwr <- system.time({
  for (j in ocean_idx[1:n_test]) {
    clm <- ts2clm(data.frame(t = dates, temp = as.numeric(sst_col[j, ])),
                  climatologyPeriod = c("1982-01-01", "2011-12-31"))
    ev <- detect_event(clm)
  }
})

per_pixel <- t_hwr[3] / n_test
n_ocean <- 6240  # pre-counted
estimated <- per_pixel * n_ocean

cat("Per pixel:      ", round(per_pixel, 3), "sec\n")
cat("Estimated total:", round(estimated), "sec (",
    round(estimated / 60, 1), "min)\n")
```

## heatwave3: single-threaded

```{r hw3-1t}
stem <- file.path(tempdir(), "bench")

t_clim <- system.time(
  ts2clm3(sst_file, name = stem,
          climatologyPeriod = c("1982-01-01", "2011-12-31"),
          var_name = "analysed_sst", n_threads = 1L, quiet = TRUE)
)

t_event <- system.time(
  detect_event3(sst_file, name = stem,
                var_name = "analysed_sst", n_threads = 1L, quiet = TRUE)
)

cat("Climatology:", round(t_clim[3], 1), "sec\n")
cat("Detection:  ", round(t_event[3], 1), "sec\n")
cat("Total:      ", round(t_clim[3] + t_event[3], 1), "sec\n")
```

## xmhw: single-thread

```{r }
# Open sst file with xarray and extract the same subset of data as the heatwaveR example above
x_ds <- xr$open_dataset(sst_file)
x_sst <- x_ds[["analysed_sst"]]$isel(
  longitude = 0L, # first longitude only (1 pixel)
  latitude  = py_slice(0L, 20L) # first 20 latitudes
)

# Check output
# print(x_sst)
# print(x_sst$shape) 

# Determine climatology data
x_clim_time <- system.time(
  x_clim <- x_threshold(x_sst)
)

# Detect events
x_event_time <- system.time(
  x_event <- x_detect(x_sst, x_clim['thresh'], x_clim['seas'])
)

cat("Climatology (estimated):", round((x_clim_time[3]/20*6240)/60, 1), "min\n")
cat("Detection (estimated):  ", round((x_event_time[3]/20*6240)/60, 1), "min\n")
cat("Total (estimated):      ", round(((x_clim_time[3]/20*6240 + x_event_time[3]/20*6240)/60), 1), "min\n")
```

## heatwave3: 12 threads

```{r hw3-12t}
t_clim_12 <- system.time(
  ts2clm3(sst_file, name = stem,
          climatologyPeriod = c("1982-01-01", "2011-12-31"),
          var_name = "analysed_sst", n_threads = 12L, quiet = TRUE)
)

t_event_12 <- system.time(
  detect_event3(sst_file, name = stem,
                var_name = "analysed_sst", n_threads = 12L, quiet = TRUE)
)

cat("Climatology:", round(t_clim_12[3], 1), "sec\n")
cat("Detection:  ", round(t_event_12[3], 1), "sec\n")
cat("Total:      ", round(t_clim_12[3] + t_event_12[3], 1), "sec\n")
```

## xmhw: 12 threads

## Results

Benchmarked on an Apple M3 Pro (12-core), macOS 15.5, R 4.6.0.

| Method | Climatology | Detection | Total | Speedup |
|--------|------------|-----------|-------|---------|
| heatwaveR (serial) | n/a | n/a | ca. 4,116 sec (68.7 min) | 1× |
| heatwave3 (1 thread) | 174 sec | 82 sec | **257 sec (4.3 min)** | **16×** |
| heatwave3 (12 threads) | 60 sec | 55 sec | **115 sec (1.9 min)** | **36×** |

The speedup comes from three sources:

1. **C++ algorithms**. Climatology (sliding-window percentile) and event
   detection (run-length encoding, gap joining, 19 metrics) are implemented
   in pure C++ rather than R + data.table.
2. **OpenMP parallelism**. Each pixel is independent, so OpenMP distributes
   pixels across cores with zero coordination overhead.
3. **Direct NetCDF I/O**. Reads the full data slab in one C call via
   libnetcdf, avoiding R's per-pixel ncdf4 random-access overhead.

### Scaling notes

- The 12-thread run is 2.2× faster than single-threaded, not 12×, because the
  full 400×200 grid must first be read into memory (ca. 900 MB for this dataset)
  and the NetCDF I/O is single-threaded. The parallelism benefit is in the
  computation phase only.
- For grids that fit in memory, I/O accounts for roughly 40–60% of wall time.
  Larger grids with more ocean pixels show better parallel scaling as the
  compute fraction grows.
- On Linux with gcc OpenMP (which has lower thread-creation overhead than macOS
  libomp), parallel scaling is typically better.
