---
title: "4D blob connectivity: heatwaves through the water column"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4D blob connectivity: heatwaves through the water column}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
has_test_data <- file.exists(
  system.file("extdata/glorys_depth_test.nc", package = "heatwave3")
)

# Chunks tagged fold_output = TRUE show their code normally but wrap the
# printed console output (e.g. ts2clm3()'s per-pixel progress messages) in a
# collapsed <details> block, since that output is long and not the point.
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  out <- default_output_hook(x, options)
  if (isTRUE(options$fold_output)) {
    sprintf("<details>\n<summary>Show console output</summary>\n\n%s\n\n</details>", out)
  } else {
    out
  }
})
```

## Overview

`detect_blob3()` finds spatially contiguous marine heatwave "blobs" by
connected-component labelling of the daily exceedance mask. Given a
depth-resolved climatology (from `ts2clm3(depth_range = ...)`, see
`vignette("depth-range")`), it now does this natively in 4D: a voxel connects
to its neighbours in longitude, latitude, depth, *and* time, instead of just
the first three. No new argument is needed -- exactly like `detect_event3()`
and `category_daily3()`, `detect_blob3()` reads the depth range straight off
`clim_file` and switches to 4D connectivity automatically.

**Connectivity is by depth *index*, not a depth *range*.** A voxel connects
to whichever level is immediately above or below it -- level 3 to level 4,
say -- regardless of how many metres apart those levels actually are. There
is no minimum/maximum vertical gap or distance threshold: two voxels one
level apart always connect if both exceed the threshold that day; two voxels
separated by an empty level in between never do, however close together in
metres those levels happen to be. This mirrors exactly how the existing
lon/lat/time connectivity already works (adjacent index, not adjacent
degrees or adjacent days-within-some-window).

This vignette uses the small 2x2 pixel, 5 depth-level GLORYS fixture bundled
with the package (`system.file("extdata/glorys_depth_test.nc", package =
"heatwave3")`). The trade-off: with
only 4 horizontal pixels, this fixture can't show interesting *spatial*
footprints. What it can show clearly is the *vertical* structure this section
adds -- a blob's footprint growing and shrinking through the water column
over its lifetime.

```{r pkgs, eval = has_test_data}
library(heatwave3)
library(ggplot2)

f <- system.file("extdata/glorys_depth_test.nc", package = "heatwave3")
```

## Detecting depth-connected blobs

First, a depth-resolved climatology covering all 5 levels in the fixture
(0.49-5.08 m):

```{r clim, eval = has_test_data, fold_output = TRUE}
stem <- file.path(tempdir(), "mhw4d")
ts2clm3(f, name = stem,
        climatologyPeriod = c("1993-01-01", "2019-12-31"),
        depth_range = c(0, 10), 
        quiet = TRUE)
```

`detect_blob3()` picks up the depth dimension from `stem_clim.nc`
automatically:

```{r blobs, eval = has_test_data}
blobs <- detect_blob3(f, paste0(stem, "_clim.nc"),
                      minVoxels = 1,
                      return = c("event", "daily", "voxel"))
nrow(blobs$event)
```

`event` and `daily` both gain depth/volume columns; `voxel` gains a `depth`
column:

```{r cols, eval = has_test_data}
names(blobs$event)
names(blobs$daily)
```

`depth_min_m`/`depth_max_m` are each blob's vertical bounding box over its
whole lifetime; `totalVolume_km3`/`peakVolume_km3`/`meanVolume_km3` are the
volumetric analogues of the existing `cumI_km2_day`/`peakArea_km2`/
`meanArea_km2` area metrics (volume = area x vertical layer thickness,
summed over the depth levels a blob occupies that day); `depthOfPeakIntensity_m`
is the depth at which the single most intense day of the blob's life
occurred. `rankBy` accepts the volume metrics too when the input is
depth-resolved:

```{r rankby, eval = has_test_data}
top <- detect_blob3(f, paste0(stem, "_clim.nc"),
                    topN = 5, rankBy = "totalVolume_km3", return = "event")
top$event[, c("blob", "rank", "duration", "depth_min_m", "depth_max_m",
             "totalVolume_km3")]
```

## How much does depth connectivity actually change?

The direct way to see the effect: for how many blobs does the vertical
bounding box span more than a single level (`depth_min_m != depth_max_m`)?
If depth connectivity did nothing, this would always be zero -- every blob
would be confined to the one level it was seeded on.

```{r fraction, eval = has_test_data}
spans_multiple <- blobs$event$depth_min_m != blobs$event$depth_max_m
mean(spans_multiple)
sum(spans_multiple)
```

On this fixture, the large majority of blobs span more than one depth
level -- voxels at adjacent levels are genuinely being unioned into shared
blobs, not kept artificially separate per level.

## Visualising a blob's vertical extent over time

Pick one blob and look at how its depth range and volume evolve day by day,
using the `daily` table:

```{r pick-blob, eval = has_test_data}
b <- top$event$blob[1]
dd <- blobs$daily[blobs$daily$blob == b, ]
range(dd$date)
```

```{r ribbon, eval = has_test_data}
ggplot(dd, aes(x = date)) +
  geom_ribbon(aes(ymin = depth_min_m, ymax = depth_max_m, fill = mean_delta)) +
  scale_y_reverse(name = "depth (m)") +
  scale_fill_viridis_c(name = expression(Delta*"T (" * degree * "C)"), option = "inferno") +
  labs(title = paste("Vertical extent of blob", b), x = NULL) +
  theme_minimal(base_size = 11)
```

Any day where the ribbon narrows (or a depth level briefly drops out
entirely) is a day the blob temporarily loses part of its vertical extent --
some level's exceedance broke the connection to the surface (or the deep
levels) that day, then reconnected once conditions there crossed the
threshold again. Note that in this example however there are no missing values.

## Depth-time structure of a single blob

The `voxel` table gives the full per-voxel footprint, which -- aggregated
across this fixture's 5 horizontal pixels since they're not the interesting
axis here -- makes a genuine depth-time "Hovmoeller" view of one blob's water
column:

```{r hovmoller, eval = has_test_data}
vox <- blobs$voxel[blobs$voxel$blob == b, ]
agg <- aggregate(delta ~ date + depth, data = vox, FUN = mean)

ggplot(agg, aes(x = date, y = depth, fill = delta)) +
  geom_tile() +
  scale_y_reverse(name = "depth (m)") +
  scale_fill_viridis_c(name = expression(Delta*"T (" * degree * "C)"), option = "inferno") +
  labs(title = paste("Depth-time structure of blob", b), x = NULL) +
  theme_minimal(base_size = 11)
```

The gaps (white tiles) are days/levels that didn't exceed the threshold that
day and so aren't part of the blob -- exactly the same days visible as
notches in the ribbon plot above, just resolved to the individual level
rather than a min/max envelope. Note however that for this example there are no gaps,
the distance between the tiles is due to the increasing size of the depth layers.

