---
title: "Depth-resolved (4D) marine heatwave analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Depth-resolved (4D) marine heatwave analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_test_data <- file.exists(
  system.file("extdata/glorys_depth_test.nc", package = "heatwave3")
)
```

## Overview

Ocean reanalyses and models such as GLORYS12 provide temperature on a 4D grid
(`lon`, `lat`, `depth`, `time`), not just at the sea surface. **heatwave3** can
compute climatologies and detect events natively across a **contiguous band
of depth levels**, producing one 4D climatology/event NetCDF instead of
per-level 3D files.

Every chunk in this vignette runs against a tiny 2x2 pixel, 5 depth-level
GLORYS12 fixture bundled with the package
(`system.file("extdata/glorys_depth_test.nc", package = "heatwave3")`) --
the same one used by the test suite -- so that this vignette actually compiles
when the documentation is built.
Everything shown works identically on a full-resolution file with many more
pixels and depth levels; just substitute your own `temp_file`. The trade-off
of the small fixture is scale, not behaviour: with only 4 horizontal pixels
and 5 depth levels (0.49-5.08 m, all within the mixed layer), don't expect
dramatic spatial or vertical structure in the figures below.

```{r setup-file, eval = has_test_data, collapse = TRUE}
library(heatwave3)

temp_file <- system.file("extdata/glorys_depth_test.nc", package = "heatwave3")
clim_period <- c("1993-01-01", "2019-12-31")
```

## Two ways to work across depth

### Option A: per-level looping

Before `depth_range` existed, the only way to get per-depth results was to
run the ordinary 3D pipeline once per level, squeezing to a single depth
index each time with the existing `depth` argument. This is still a
legitimate, simple approach: it can be run in parallel across depth and
reuses the same validated 3D code path as every other heatwave3 analysis.

```{r per-level-loop, eval = has_test_data, quiet = TRUE}
n_depth <- 5L  # number of depth levels in this fixture

depth_indices <- seq_len(n_depth) - 1L  # depth= is a 0-based index: 0, 1, 2, 3, 4
purrr::walk(depth_indices, \(k) {
  detect3(temp_file, name = file.path(tempdir(), paste0("mhw_z", k)),
          climatologyPeriod = clim_period,
          depth = k, category = TRUE, quiet = TRUE)
})
```

The trade-off: this produces `n_depth` separate pairs of `_clim.nc`/
`_events.nc` files, each with `n_depth`-fold repeated I/O over the same
lon/lat/time grid, and there is no vertical linkage between levels (each is
detected completely independently, as if it were its own 3D dataset). For
users who just want "climatology and events at 20 m, 50 m, 100 m..." this may
already be enough -- stack the files yourself downstream with `terra` (R),
 `xarray` (python), or `CDO` (bash).

### Option B: native `depth_range` (recommended)

`depth_range = c(min_depth, max_depth)` (metres, inclusive) tells `ts2clm3()`
and `detect3()` to read every depth level whose coordinate falls in that
range and compute an independent climatology/event table for every
`(lon, lat, depth)` triple, in **one** call producing **one** 4D file. This
avoids the repeated I/O and per-level file bookkeeping of Option A above. The
fixture's 5 levels all fall within 0-10 m:

```{r ts2clm-depth-range, eval = has_test_data, collapse = TRUE}
stem <- file.path(tempdir(), "mhw4d")
clim_file <- ts2clm3(file_in = temp_file,
                     name = stem,
                     climatologyPeriod = clim_period,
                     depth_range = c(0, 10),
                     n_threads = 2)
```

`detect_event3()` needs no separate depth argument at all: it reads the depth
range straight back out of the climatology file it is already loading, so
climatology and event detection can never end up out of sync about which
levels they are comparing.

```{r detect-event-depth-range, eval = has_test_data, collapse = TRUE}
event_file <- detect_event3(file_in = temp_file,
                            name = stem,
                            category = TRUE,
                            n_threads = 2, 
                            quiet = TRUE)
```

Or as a single call via `detect3()`:

```{r detect3-depth-range, eval = has_test_data, collapse = TRUE}
detect3(file_in = temp_file,
        name = stem,
        climatologyPeriod = clim_period,
        depth_range = c(0, 10),
        category = TRUE,
        n_threads = 2)
```

## Reading the depth-resolved output

`hw3_export()` surfaces a `depth` column automatically whenever the source
file is depth-resolved:

```{r export-depth, eval = has_test_data}
clim_df  <- hw3_export(paste0(stem, "_clim.nc"))
event_df <- hw3_export(paste0(stem, "_events.nc"))

head(clim_df)
table(round(event_df$depth, 2))
```

Because every `(lon, lat, depth)` triple gets its own climatology, the
seasonal cycle and threshold both flatten out with depth -- the surface has
the largest seasonal amplitude and the highest threshold, decaying as you
move down through the mixed layer. This is already visible directly from the
exported climatology data frame, with no new plotting functions required
(the effect is subtle here since all 5 levels sit within the top 5 m; it's
more pronounced across a deeper range on a real file):

```{r plot-clim-by-depth, eval = has_test_data}
library(ggplot2)

one_pixel <- subset(clim_df, lon == clim_df$lon[1] & lat == clim_df$lat[1])

ggplot(one_pixel, aes(doy, seas, colour = factor(round(depth, 1)))) +
  geom_line(linewidth = 0.7) +
  scale_colour_viridis_d(name = "depth (m)") +
  labs(title = "Seasonal climatology by depth", x = "Day of year", y = "degC") +
  theme_minimal(base_size = 11)
```

A depth-time Hovmoeller-style view of event intensity is also already
possible by building the plot directly from `event_df`, without any new
graphics functions:

```{r plot-hovmoller, eval = has_test_data}
one_pixel_events <- subset(event_df, lon == event_df$lon[1] & lat == event_df$lat[1])

ggplot(one_pixel_events, aes(x = date_peak, y = depth, fill = intensity_max)) +
  geom_tile(height = 1) +
  scale_fill_viridis_c(name = expression(I[max] ~ "(degC)"), option = "inferno") +
  scale_y_reverse(name = "depth (m)") +
  labs(title = "Event intensity by depth and peak date", x = NULL) +
  theme_minimal(base_size = 11)
```

## Depth-resolved event-line plots with `geom_flame3()` / `geom_lolli3()`

The per-day product also carries a depth dimension now, so `daily = "also"`
works together with `depth_range` -- no separate depth argument needed here
either, for the same reason `detect_event3()` doesn't need one:

```{r detect3-daily-depth, eval = has_test_data, collapse = TRUE}
detect3(file_in = temp_file,
        name = stem,
        climatologyPeriod = clim_period,
        depth_range = c(0, 10),
        category = TRUE,
        daily = "also",
        n_threads = 2)
```

`hw3_export()` surfaces the same `depth` column on this per-day product, and
its streaming subset (`lon_range`/`lat_range`/`time_range`/`depth_range`/`n`)
is depth-aware too, so a single pixel's full depth profile over a short date
window reads cheaply without loading the whole file:

```{r export-daily-depth, eval = has_test_data}
daily_df <- hw3_export(paste0(stem, "_events_daily.nc"),
                       lon_range = c(25.16, 25.17), lat_range = c(-34.92, -34.91),
                       time_range = c("1999-01-01", "2000-12-31"))
head(daily_df)
```

`geom_flame3()` automatically folds a mapped `depth` aesthetic into its row
grouping -- each depth level gets its own exceedance runs instead of being
treated as one continuous series with depths interleaved -- so one call plots
every level:

```{r geom-flame-depth, eval = has_test_data, fig.height = 9}
ggplot(daily_df, aes(x = t, y = temp, y2 = thresh, depth = depth)) +
  geom_flame3() +
  geom_line(linewidth = 0.3) +
  facet_wrap(~ depth, ncol = 1, labeller = label_both) +
  labs(title = "Depth-resolved event-line plot", x = NULL,
       y = expression("Temperature (" * degree * "C)")) +
  theme_minimal(base_size = 11)
```

`event_line3()` also gained a `depth` argument, for the classic single-pixel
flame plot at one chosen level (matched to the nearest depth actually present
in the climatology, here 2.6 m):

```{r event-line-depth, eval = has_test_data}
event_line3(temp_file, paste0(stem, "_clim.nc"),
            lon = 25.16667, lat = -34.91667, depth = 3,
            event_file = paste0(stem, "_events.nc"))
```

`geom_lolli3()` needs no special handling for depth: each lollipop is drawn
from its own row independently of any grouping, so map `depth` to colour (or
facet) like any other column:

```{r geom-lolli-depth, eval = has_test_data}
event_df_1px <- hw3_export(paste0(stem, "_events.nc"),
                           lon_range = c(25.16, 25.17), lat_range = c(-34.92, -34.91))

ggplot(event_df_1px, aes(x = date_peak, y = intensity_max, colour = factor(round(depth, 1)))) +
  geom_lolli3() +
  scale_colour_viridis_d(name = "depth (m)") +
  labs(title = expression(I[max] ~ "by depth and peak date (single pixel)"),
       x = NULL, y = expression(I[max] ~ "(degC)")) +
  theme_minimal(base_size = 11)
```

