---
title: "NetCDF output internals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NetCDF output internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

heatwave3 produces two types of NetCDF output files:

1. **Climatology file** (from `ts2clm3()`), holding seasonal and threshold
   climatologies per pixel per day-of-year
2. **Event file** (from `detect_event3()`), holding detected events stored as a
   CF ragged array

Both files target CF-1.8 compliance and can be read by any NetCDF-capable tool
(R ncdf4, Python xarray/netCDF4, CDO, ncdump, Panoply).

## Climatology file structure

Produced by `ts2clm3()`. Contains the 366-day climatological cycle for each
pixel.

### Dimensions

| Dimension | Size | Description |
|-----------|------|-------------|
| `lon`     | nlon | Longitude grid points |
| `lat`     | nlat | Latitude grid points |
| `doy`     | 366  | Day of year (1–366, including leap day) |

### Variables

| Variable | Dimensions | Units | Description |
|----------|-----------|-------|-------------|
| `lon`    | (lon)     | degrees_east | Longitude coordinate |
| `lat`    | (lat)     | degrees_north | Latitude coordinate |
| `doy`    | (doy)     | none | Day of year index |
| `seas`   | (lon, lat, doy) | degC | Seasonal mean climatology |
| `thresh` | (lon, lat, doy) | degC | Threshold climatology (such as the 90th percentile) |
| `var`    | (lon, lat, doy) | degC | Variance climatology (optional, if `var = TRUE`) |

### Global attributes

| Attribute | Example | Description |
|-----------|---------|-------------|
| `Conventions` | CF-1.8 | CF compliance level |
| `created_by` | heatwave3 | Package identifier |
| `source_file` | /path/to/sst.nc | Input SST file |
| `climatologyPeriod_start` | 1982-01-01 | Baseline start |
| `climatologyPeriod_end` | 2011-12-31 | Baseline end |
| `pctile` | 90 | Percentile used for threshold |
| `windowHalfWidth` | 5 | Sliding window half-width |
| `smoothPercentileWidth` | 31 | Rolling mean smoothing width |

### Reading in R

```{r read-clim}
library(ncdf4)
nc <- nc_open("clim_output.nc")

lon <- ncvar_get(nc, "lon")
lat <- ncvar_get(nc, "lat")
seas <- ncvar_get(nc, "seas")    # [doy, lat, lon] in R
thresh <- ncvar_get(nc, "thresh")

# Extract DOY 1 map
seas_jan1 <- seas[1, , ]

nc_close(nc)
```

### Reading in Python

```python
import xarray as xr

ds = xr.open_dataset("clim_output.nc")
ds["seas"].sel(doy=1).plot()
```

### Reading with CDO

```bash
# Extract seasonal climatology as a time series (366 steps)
cdo selvar,seas clim_output.nc seas_only.nc
```

## Event file structure

Produced by `detect_event3()`. Uses a **flat event dimension**, where each row
represents one detected event at one pixel. This avoids padding to a maximum
number of events per pixel and keeps the file compact.

### Dimensions

| Dimension | Size | Description |
|-----------|------|-------------|
| `event`   | N    | Total number of events across all pixels |

### Coordinate variables

| Variable | Type | Description |
|----------|------|-------------|
| `lon`    | double | Longitude of the pixel where the event occurred |
| `lat`    | double | Latitude of the pixel |
| `pixel_index` | int | Linear pixel index (ilon × nlat + ilat) |

### Event metric variables

All have dimension `(event)`:

| Variable | Units | Description |
|----------|-------|-------------|
| `event_no` | none | Event number within the pixel |
| `date_start` | days since ref | Start date |
| `date_peak` | days since ref | Peak date |
| `date_end` | days since ref | End date |
| `duration` | days | Event duration |
| `intensity_mean` | degC | Mean intensity (rel. to seasonal climatology) |
| `intensity_max` | degC | Maximum (peak) intensity |
| `intensity_var` | degC | Intensity variability (std. dev.) |
| `intensity_cumulative` | degC days | Cumulative intensity |
| `intensity_mean_relThresh` | degC | Mean intensity (rel. to threshold) |
| `intensity_max_relThresh` | degC | Max intensity (rel. to threshold) |
| `intensity_var_relThresh` | degC | Variability (rel. to threshold) |
| `intensity_cumulative_relThresh` | degC days | Cumulative (rel. to threshold) |
| `intensity_mean_abs` | degC | Mean absolute temperature |
| `intensity_max_abs` | degC | Max absolute temperature |
| `intensity_var_abs` | degC | Absolute temperature variability |
| `intensity_cumulative_abs` | degC days | Cumulative absolute temperature |
| `rate_onset` | degC/day | Rate of onset |
| `rate_decline` | degC/day | Rate of decline |

### Date encoding

Date fields (`date_start`, `date_peak`, `date_end`) are stored as integer
offsets in units of `"days since YYYY-MM-DD"`, where the reference date is the
first time step of the input SST file. To convert to actual dates:

```{r read-events}
library(ncdf4)
nc <- nc_open("event_output.nc")

# Read date_start and convert to Date
ds <- ncvar_get(nc, "date_start")
ds_units <- ncatt_get(nc, "date_start", "units")$value
ref_date <- as.Date(sub("days since ", "", ds_units))
start_dates <- ref_date + ds

# Read event metrics
i_max <- ncvar_get(nc, "intensity_max")
duration <- ncvar_get(nc, "duration")
event_lon <- ncvar_get(nc, "lon")
event_lat <- ncvar_get(nc, "lat")

nc_close(nc)
```

### Global attributes

| Attribute | Description |
|-----------|-------------|
| `Conventions` | CF-1.8 |
| `source_file` | Path to the input SST file |
| `climatology_file` | Path to the climatology file used |
| `minDuration` | Minimum event duration setting |
| `maxGap` | Maximum gap setting |
| `coldSpells` | 0 or 1 |

## CF compliance notes

heatwave3 output files aim for CF-1.8 compliance:

**Compliant:**

- `Conventions` global attribute set to "CF-1.8"
- Coordinate variables have `units` (`degrees_east`, `degrees_north`)
- Coordinate variables have `axis` attributes (X, Y, T)
- Data variables have `long_name` and `units`
- Date variables use CF time encoding (`days since ...`)
- NetCDF-4 format with compression (deflate level 4)

**Best-effort (not strictly required by CF):**

- The event file uses a flat dimension rather than CF indexed ragged arrays
  (`sample_dimension` / `instance_dimension` attributes are not set). This is a
  pragmatic choice, since the flat layout is simpler to read and write, and all
  major tools handle it correctly.
- `standard_name` attributes are set on coordinate variables but not on all
  data variables (the 19 event metrics do not have CF standard names, as MHW
  metrics are domain-specific).

## Alternative output formats

In addition to NetCDF, heatwave3 can export output as CSV, RDA, or Parquet
using `hw3_export()` or the `save_format` argument:

```{r export-formats}
# During detection
detect_event3(file_in, clim_file, event_file, save_format = "parquet")

# After the fact
hw3_export("event_output.nc", format = "csv", type = "event")
hw3_export("clim_output.nc", format = "rda", type = "clim")
```

**Format comparison:**

| Format | Size | Speed | Interop | Notes |
|--------|------|-------|---------|-------|
| NetCDF | Small (compressed) | Fast (C library) | Excellent | Primary format, always produced |
| Parquet | Small (columnar) | Fast | Good (R/Python/Spark) | Requires `arrow` package |
| RDA | Medium | Fast (R native) | R only | Contains a list named `hw3_data` |
| CSV | Large (text) | Slow | Universal | Not recommended for large grids |
