---
title: "Parallel performance: in-process threads vs R-side parallelism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel performance: in-process threads vs R-side parallelism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Two kinds of parallelism

`heatwave3` detects marine heatwaves one pixel at a time, and pixels are
completely independent of one another: a pixel's climatology and events depend
only on its own temperature time series. That independence is what makes the
problem *embarrassingly parallel*, and it can be exploited in two distinct
ways.

1. **In-process threads.** The C++ core spreads pixels across CPU cores using
   `std::thread`. Everything happens inside a single R process: the data slab is
   read once into shared memory, the threads compute, and one set of output
   files is written. Controlled by the `n_threads` argument (or
   `setHW3threads()`). This is the fast path on a single machine, and it works
   out of the box on every platform with no OpenMP toolchain required.

2. **R-side parallelism (across processes).** Several R processes each run
   `heatwave3` single-threaded on a *spatial tile* of the grid, and the results
   are combined. Controlled with the `parallel` package. It is the only way to
   scale across more than one machine.

This vignette measures both, on the same dataset, and explains when to reach
for each.

> **Backend note.** Earlier versions used OpenMP for the in-process threads.
> heatwave3 now uses `std::thread`, which needs no OpenMP runtime (no `libomp`,
> no `omp.h`, no special install step) and avoids the macOS `libomp` worker-thread
> crash that struck heavily-loaded sessions (terra/GDAL, conda toolchains, or the
> Positron/`ark` kernel). The [Results](#results) below compare the two backends.

## Is multithreading active?

`getHW3threads()` reports the default thread count (roughly half your cores).
`n_threads` is honoured on every platform; there is no build-time switch to
miss.

```{r check-omp}
library(heatwave3)
getHW3threads()
#> [1] 9        # default: 50% of 18 cores
```

The shared library links no OpenMP runtime; it uses `std::thread` (pthreads):

```sh
otool -L $(Rscript -e 'cat(system.file("libs/heatwave3.so", package="heatwave3"))')   # macOS
ldd       $(Rscript -e 'cat(system.file("libs/heatwave3.so", package="heatwave3"))')   # Linux
```

## The dataset

The OSTIA reanalysis over the South-West Indian Ocean / South African east
coast: a 400 × 200 grid at 0.05° resolution (15–35°E, 38–28°S), 14,276 daily
time steps (1982–2021), of which roughly 50,000 pixels are ocean. The
climatological baseline is 1982–2011.

```{r data}
sst_file <- "/Volumes/OceanData/OSTIA_East_Coast_MHW/SWIO_Jan1982-Dec2021.nc"
var_name <- "analysed_sst"
clim_period <- c("1982-01-01", "2011-12-31")
```

All three methods produce the **same deliverable**, the full event table as an
in-memory `data.frame`, so their wall-clock times are directly comparable.

## Method 1: single-threaded

The baseline. `n_threads = 1` forces one core regardless of the package
default.

```{r single}
library(heatwave3)

run_once <- function(nthreads, lon_range = NULL) {
  stem <- tempfile()
  on.exit(unlink(paste0(stem, c("_clim.nc", "_events.nc"))))
  out <- detect3(
    sst_file,
    name = stem,
    climatologyPeriod = clim_period,
    var_name = var_name,
    lon_range = lon_range,
    category = TRUE,
    n_threads = nthreads,
    quiet = TRUE
  )
  hw3_export(out[["events"]])
}

system.time(ev1 <- run_once(1L))
```

## Method 2: in-process multithreading

Identical call, more threads. Because pixels are independent, each thread takes
a share of them, with no locking and no coordination.

```{r openmp}
system.time(ev12 <- run_once(12L))

# Results are independent of thread count:
nrow(ev1) == nrow(ev12)
#> [1] TRUE
```

Sweeping the thread count maps the scaling curve:

```{r openmp-sweep}
for (nt in c(1, 2, 4, 8, 12, 16)) {
  t <- system.time(run_once(nt))[["elapsed"]]
  cat(sprintf("%2d threads: %5.1f s\n", nt, t))
}
```

## Method 3: R-side parallelism (PSOCK)

To scale across machines, or to parallelise the read and the assembly as well
as the compute, split the grid into longitude tiles and run one single-threaded
`heatwave3` per worker process. Since pixels are independent, the union of the
tiles is **identical** to a whole-domain run.

We use a **PSOCK cluster** (independent R processes) rather than fork-based
`mclapply()`. On macOS, forking after the NetCDF/HDF5 and Apple Objective-C
runtimes have initialised is unsafe (sporadic worker crashes, and the
`+[NSCharacterSet initialize] ... fork()` abort). PSOCK launches clean
processes that each open their own files, like running several `Rscript` jobs
side by side. It is also the only option on Windows, where `fork()` does not
exist.

```{r psock}
library(parallel)

# Read the longitude coordinate once (cheap), to cut non-overlapping tiles.
g <- heatwave3:::hw3_read_sst(
  sst_file,
  var_name,
  time_range = c("1982-01-01", "1982-01-02")
)
lon <- sort(unique(g$lon))

make_tiles <- function(n) {
  pad <- 0.45 * min(diff(lon)) # < half the grid spacing
  grp <- cut(seq_along(lon), breaks = n, labels = FALSE)
  unname(lapply(split(lon, grp), function(L) c(min(L) - pad, max(L) + pad)))
}

run_psock <- function(n_workers) {
  cl <- makeCluster(n_workers, setup_strategy = "sequential")
  on.exit(stopCluster(cl))
  clusterEvalQ(cl, suppressPackageStartupMessages(library(heatwave3)))
  clusterExport(cl, c("sst_file", "var_name", "clim_period"))
  tiles <- make_tiles(n_workers)
  parts <- parLapply(cl, tiles, function(lr) {
    stem <- tempfile()
    on.exit(unlink(paste0(stem, c("_clim.nc", "_events.nc"))))
    out <- detect3(
      sst_file,
      name = stem,
      climatologyPeriod = clim_period,
      var_name = var_name,
      lon_range = lr,
      category = TRUE,
      n_threads = 1L,  # 1 thread per worker!
      quiet = TRUE
    )
    hw3_export(out[["events"]])
  })
  do.call(rbind, parts)
}

system.time(evp <- run_psock(16L))
```

The `n_threads = 1L` inside each worker is essential: each freshly-loaded
`heatwave3` otherwise defaults to *half the cores*, so 16 workers each spawning
9 threads would oversubscribe the machine badly. See
[Don't oversubscribe](#dont-oversubscribe).

## Results

```{r results-data, eval = TRUE, echo = FALSE}
# Wall-clock seconds to build the full event data.frame for the whole 400x200
# domain (4,863,985 events; identical for every run). macOS (Darwin 25.5, Apple
# Silicon arm64), 18 cores, R 4.5.3, warm OS file cache. The OpenMP figures are
# from the previous release (2026-06-02); the std::thread and PSOCK figures were
# measured on the same machine after the backend switch (2026-06-04).
# Source: dev/benchmark_threads.R.
perf <- data.frame(
  method  = c(rep("stdthread", 6), rep("openmp", 6), rep("psock", 3)),
  n       = c(1, 2, 4, 8, 12, 16,   1, 2, 4, 8, 12, 16,   4, 8, 16),
  elapsed = c(112.03, 73.54, 52.73, 42.07, 38.04, 37.26,   # std::thread (now)
              109.30, 72.47, 50.28, 40.16, 36.95, 35.67,   # OpenMP (previous)
               44.36, 25.52, 18.36)                        # PSOCK
)
```

```{r results-table, eval = TRUE, echo = FALSE, results = "asis"}
if (nrow(perf)) {
  base <- perf$elapsed[perf$method == "stdthread" & perf$n == 1]
  perf$speedup <- round(base / perf$elapsed, 1)
  lab <- c(stdthread = "Threads (std::thread, now)",
           openmp    = "Threads (OpenMP, previous)",
           psock     = "PSOCK (R workers)")
  ord <- c("stdthread", "openmp", "psock")
  tab <- perf[order(match(perf$method, ord), perf$n), ]
  knitr::kable(
    data.frame(
      Method = lab[tab$method],
      `Workers/threads` = tab$n,
      `Time (s)` = round(tab$elapsed, 1),
      `Speed-up` = tab$speedup,
      check.names = FALSE
    ),
    row.names = FALSE,
    caption = paste("Wall-clock time to produce the full event table.",
                    "Speed-up is relative to the single-threaded baseline.")
  )
}
```

```{r results-plot, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 4.5, fig.alt = "Speed-up versus thread or worker count for std::thread, the previous OpenMP build, and PSOCK, against the ideal linear line. The std::thread and OpenMP thread curves nearly overlap; PSOCK rises higher."}
if (nrow(perf)) {
  library(ggplot2)
  base <- perf$elapsed[perf$method == "stdthread" & perf$n == 1]
  perf$speedup <- base / perf$elapsed
  ideal <- data.frame(n = seq(1, max(perf$n)))
  labs_v <- c(stdthread = "std::thread (now)",
              openmp = "OpenMP (previous)", psock = "PSOCK")
  ggplot(perf, aes(n, speedup, colour = method, linetype = method)) +
    geom_line(
      data = ideal,
      aes(n, n),
      inherit.aes = FALSE,
      linetype = 2,
      colour = "grey60"
    ) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    annotate(
      "text",
      x = max(perf$n) * 0.78,
      y = max(perf$n) * 0.92,
      label = "ideal linear",
      colour = "grey55",
      size = 3.3
    ) +
    scale_colour_manual(
      values = c(stdthread = "#1b9e77", openmp = "#7570b3", psock = "#d95f02"),
      labels = labs_v, name = NULL
    ) +
    scale_linetype_manual(
      values = c(stdthread = "solid", openmp = "dashed", psock = "solid"),
      labels = labs_v, name = NULL
    ) +
    labs(
      x = "threads / worker processes",
      y = "speed-up vs 1 thread",
      title = "Parallel scaling: std::thread vs OpenMP vs R-side (PSOCK)"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top")
}
```

On this run (whole domain, ca. 50,000 ocean pixels, warm file cache), the
single-threaded baseline took **112 s**. In-process `std::thread` reached
**3.0×** (37 s) at 16 threads, and R-side PSOCK reached **6.1×** (18 s) at 16
workers. Every run returned the same 4,863,985 events.

### std::thread vs OpenMP

This release replaced the OpenMP backend with C++ `std::thread`. The switch was
performance-neutral: at matched thread counts `std::thread` ran within about
5% of the previous OpenMP build, tracking the same scaling curve. At 16
threads it took 37.3 s against OpenMP's 35.7 s; at 4 threads, 52.7 s against
50.3 s. That small, consistent overhead buys two things, namely freedom from the
macOS `libomp` worker-thread crash that struck when heatwave3 ran alongside
heavy dynamic libraries (terra, sf, GDAL) in the same process, and the removal
of the OpenMP build dependency altogether. The dashed line in the figure is the
old OpenMP build; the solid green line, the current `std::thread` build, sits
just beneath it.

### Reading the curves

Both threading and PSOCK speed the analysis up substantially, and both fall
short of ideal linear scaling, but PSOCK scales further, for a reason worth
understanding.

The work has three phases, namely **read** the NetCDF slab, **compute** per
pixel, and **assemble** the 4.9-million-row event `data.frame`. In-process
threads parallelise only the middle phase. The read and the assembly happen
single-threaded in the one R process, so by Amdahl's law that serial fraction
caps the speed-up, and the thread curve plateaus near 3× regardless of thread
count.

PSOCK parallelises **all three** phases. Each worker process reads only its own
longitude slice and builds only its own sub-frame, concurrently, so the I/O and
the assembly scale with the worker count too. With the file already in the OS
cache, those concurrent reads are cheap, and PSOCK keeps gaining past the point
where in-process threading has flattened.

Two caveats keep this from being a blanket "PSOCK wins":

- **Endpoint.** This benchmark reads the events back into an in-memory
  `data.frame` (via `hw3_export()`), which makes the serial assembly a large
  part of the thread build's tail. The usual production endpoint (just write the
  event NetCDF and stop) removes that assembly, shrinking the serial fraction and
  narrowing the gap.
- **Storage.** PSOCK's edge depends on those per-worker reads being cheap. On a
  cold cache or slow/network disk, sixteen processes reading at once can
  contend, eroding or reversing the advantage. The in-process threads read the
  slab exactly once.

In-process threading also stays the *simpler* option, with one call, one output
file, no tiling or merge, and no inter-process serialisation. PSOCK earns its
overhead by parallelising the read and the assembly as well as the compute, by
isolating failures to a single tile, and by scaling across separate machines.

## Discussion

### When to use which

- **In-process threads (`std::thread`)** are the default on a single machine.
  They have the lowest overhead, use shared memory, and write one NetCDF. Set
  `n_threads`, or leave it at the package default. Nothing to build or enable.
- **R-side parallelism (PSOCK)** is for when (a) you want the read and the
  assembly parallelised too, not just the per-pixel compute, (b) you want to
  spread work across **multiple machines** in a cluster, or (c) you want process
  isolation so one failing tile cannot bring down the whole run. Each worker
  writes its own tile, and you merge afterwards.
- The two combine across a cluster. One PSOCK (or MPI) worker per node, each
  using `n_threads` across that node's cores, is the standard HPC pattern. On a
  single workstation, pick one.

### Correctness

Because pixels are independent, every method returns the **identical** event
table (we assert `nrow()` equality above, and the per-pixel metrics match to
the bit). Tiling changes only *which* process computes a pixel, never the
result. The one exception is `detect_blob3()`, which finds spatially
**connected** heatwave blobs across pixels. That computation must see the whole
grid and **cannot** be tiled this way.

### Do not oversubscribe {#dont-oversubscribe}

Running *N* worker processes that each launch *M* threads creates
*N × M* threads. If that exceeds your physical cores, the threads contend for
them and everything slows down. Practical limits:

- Pure threads: `n_threads` ≤ physical cores.
- Pure PSOCK: `n_workers` ≤ physical cores, **`n_threads = 1` in every worker**.
- Hybrid (cluster): `n_workers_per_node × n_threads ≈ cores_per_node`.

`heatwave3` defaults to ca. 50% of cores so that a careless nested call does
not saturate the machine, but explicit is better than implicit.

## Platform notes

In-process multithreading is built in via `std::thread`, so there is **no
OpenMP toolchain to install or enable** -- no `libomp`, no `omp.h`, no
build-time probe. `n_threads` works on every platform once the package builds.
The only external dependency is the NetCDF C library; `configure` locates it and
reports:

```
Using netcdf:
  CPPFLAGS: -I/opt/homebrew/include
  LIBS:     -L/opt/homebrew/lib -lnetcdf
Parallelism: C++ std::thread (no OpenMP runtime required)
```

### Linux

```sh
sudo apt install libnetcdf-dev build-essential   # Debian/Ubuntu
sudo dnf install netcdf-devel gcc-c++             # Fedora/RHEL
R CMD INSTALL heatwave3
```

Fork-based `mclapply()` is also safe here, though PSOCK remains the portable
choice for R-side parallelism.

### macOS

```sh
brew install netcdf      # the only external requirement
R CMD INSTALL heatwave3
```

`std::thread` needs no Homebrew `libomp` and no `omp.h`. It also removes a
class of crash seen on earlier OpenMP builds: a `dlopen`-ed `libomp` could fail
to allocate worker-thread storage in a process with a large native footprint
(`terra`/`raster`/`sf` and their GDAL stack, conda toolchains, or the
Positron/`ark` kernel that embeds R) and segfault in `__kmp_suspend`. With
`std::thread` that cannot happen, regardless of load order or front-end.

For **R-side** parallelism on macOS, prefer PSOCK. If you must use `mclapply()`,
export `OBJC_DISABLE_INITIALIZE_FORK_SAFETY=YES` before starting R, and treat
concurrent forked NetCDF/HDF5 access as unsafe.

### Windows

`configure.win` locates NetCDF and multithreading via `n_threads` works with no
extra steps. There is no `fork()` on Windows, so `mclapply()` silently runs
serially, and R-side parallelism **must** use a PSOCK cluster (`makeCluster()` /
`parLapply()`), exactly as shown above.

## Summary

| Approach | Enable with | Best when | Notes |
|----------|-------------|-----------|-------|
| Single-threaded | `n_threads = 1` | debugging, tiny grids | reference result |
| Threads (`std::thread`) | `n_threads = N` | one machine | low overhead, shared memory, single output, no OpenMP needed |
| PSOCK | `parallel::makeCluster` | many machines, or process isolation | tile by longitude, `n_threads = 1` per worker |

All three give the same science. In-process threads are the efficient default on
a single host, and R-side parallelism is the route to multi-machine scaling and
process isolation.
